A Comprehensive Analysis Revealing Contributing Factors Behind Declining US Birth Rates

### 2024-02-15 paper/paper.qmd
fixes and tweaks and I just wanna sleep

---
title: "A Comprehensive Analysis Revealing Contributing Factors Behind Declining US Birth Rates"
#subtitle: "My subtitle if needed"
author: Maroosh Gillani, Aaron Liu & Jimmy Luc
thanks: "Code and data are available at: https://github.com/rex009x/us_birth_rates/tree/master/data. Selected aspects of the original paper that underpins this paper were reproduced and a record of this is available at: https://doi.org/10.48152/ssrp-nsc2-4v21 "
date: today
date-format: long
abstract: "The United States has been experiencing a steady decline in their birth rates, as suggested by Kearney, Levine, and Pardue's (2020) article 'The Puzzle of Falling US Birth Rates since the Great Recession'. Through the analysis of multiple different factors, such as age and education, we notice a continual decline in birth rate starting from the year 2007. We discover how different social, cultural, and economic shifts play a part in this worrying trend. This paper demonstrates how socioeconomic trends have affected Americans to shift their priorities, and how this change represented through data has played a key role in the irreversible downward trajectory of US birth rates."
format:
  pdf:
    fig-pos: "H"
toc: true
number-sections: true
bibliography: references.bib
---
```{r, include=FALSE, warning=FALSE, message=FALSE}
# workplace setup
library(tidyverse)
library(dplyr)
library(ggplot2)
library(here)
library(haven)

@@

The nation's birth rate has fallen significantly since 2007, dropping about 23% by 2022, measured by the number of births per 1000 women [@citeAxios] and that birth rates tend to fall based on the economy. 2007 marks the start of the most severe economic crisis the world has ever faced, spanning the years 2007 until 2009. The events of this crisis hve had long lasting damage on the economy, with unemployment doubling and the crash of the housing market [@citeFDR]. These poor economic conditions and the unusually slow recovery of the economy would play a large role in the decision to have children, as an unpredictable economic environment makes people feel less secure about raising a family, thus contributing to the recorded decline in birth rates. This is just one of various factors that we will be exploring in this paper to observe any trends and correlations between societal and economic shifts and the dwindling birth rate in the US.

@@

The data for this paper was acquired through the replication package of Kearney, Levine, and Pardue's (2020) article "The Puzzle of Falling US Birth Rates since the Great Recession" available through the American Economic Association (AEA), with the aid of data available from Centers for Disease Control, and Prevention (CDC) Natality data [@citeCDCNCHS] and CDC SEER U.S. Population Data [@citeCDCSeer]. The data sources were accessed on February 12th, 2024 and acts as the foundation of this paper. The data sources themselves are a collaborative effort including but not limited to the U.S. Census Bureau, the National Cancer Institute, SEER Data, CDC, and the National Center for Health Statistics (NCHS). The United States census is legally mandated by the United States Constitution which occurs every 10 years. The data consists largely of `numeric` values representing years, birth rates by race and ethnicity, marital status, education, age, and birth parity, or `character` values representing US states.

All data collection and analysis was done using statistical tools and data visualization program R [@citeR] and Rstudio IDE [@citeRStudio] to help streamline the workflow. The analysis was done using the R program and the following supporting packages `tidyverse` [@citeTidyverse], `haven` [@citeHaven], `ggplot2` [@citeGgplot2], `here` [@citeHere], `ggthemes` [@citeGgthemes], and `dplyr` [@citeDplyr]. We will provide more comprehensive details about the data cleaning, and analysis in the following subsections.

@@

The overarching variable in our measurements is birth rate. Birth rate is measured by the number of births per 1000 women. Using this metric, we graph it in the context of the other variables included in our research we stated above.

@@

The data we acquired through various sources and methods as previously mentioned presents a plethora of data points which we concern ourselves with. In order to produce our desired visualizations, we must first begin by processing our data in preparation. While a portion of the data has already been processed by the authors of "The Puzzle of Falling US Birth Rates since the Great Recession" [@citeOriginalReplication], the remaining data sets are abundant. Data that can be so specific such as the population of women the age of 15 who live in Arkansas.

With the consideration of our variables of interest in mind, we can proceed with the processing of the data. As birth rates are the main focal point of this paper, we are interested in the number of births that occur throughout the United States and the total population at any given time frame. The data including US Census Data and NCHS Vital Statistics Natality Birth Data provides this information by listing out the number of women and child births per state by year. Together, we have the sufficient information required to proceed with the visualization for the number of births per 1000 women in the US.

Next, we are interested to find the number of births per 1000 women in regards to the woman's age. As previously mentioned, we have the required data points concerning the population of women but now incorporating another variable of interest being the age. Proceeding, we intend to organize the data into age groups in order to observe the potential generational trends in birth rate. Fortunately, the data is already prepared containing the birth rates per year for each corresponding five-year age group [@citeOriginalReplication].

@@

Considering births by "parity", referring to the number of children for a given woman, we are interested in the data containing information regarding the number of first, second, third, and fourth+ births in the US. Combining NCHS Natality Microdata and CDC SEER population data, and simply filter for the data points of interest to receive the information required to generate our visualizations.

Similarly to the data processing to acquire the birth rates per year by five-year age groups, gathering NCHS Vital Statistics birth data and CDC SEER population data we organize the data by mother's age and birth cohorts. Data processing and visualization of @fig-3 was completed with the aid of Kimlin Chin's replication [@citeAnotherReplication] of Kearney, Levine, and Pardue's (2020) article "The Puzzle of Falling US Birth Rates since the Great Recession".

@@

The measurement of birth rates is nothing short of a difficult task upon consideration of all its nuances. What exactly defines a birth and what parameters determine how it is recorded? This variable of interest becomes slightly cloudy when answering these questions. Do births that occur outside of a hospital count to the total number of births? The bounds of what a birth is becomes even blurrier when we consider miscarriages and related factors.

@@

Measuring the level of education appears to be simple but is more inconsistent than initial thought. Data acquired through the CPS which is a monthly voluntary survey where U.S. households are selected [@citeCPS]. This is important to note as there may be traces of response bias and other forms of biases. As a result, it is difficult to filter the data from these bias entries and the generated visualizations may be skewed.

The number of children for a given mother or "parity" is a rather straight-forward measurement. Although questions concerning whether legal adoption of a child contributes to this measurement, it is unclear to find an answer. A more concise explanation for the properties of this measurement would greatly benefit the reader's experience.

@@

The most likely explanation that acts as the root cause for the decline in US births is the shift in priorities across cohorts of women [@citeEconomicPolicy]. While there are many who can speculate and point fingers at specific events that occurred pre- or post-2007 such as the "inevitable" financial crisis more commonly known as "The Great Recession". Fertility is procyclical: there exists a relationship between the current economic state and fertility rate [@citeSageJournals]. This alone would suffice as the reasoning for such drop in fertility but instead the shifting priorities of young adults adopting a new and modern take on life is the overarching explanation to the downward trajectory of US birth rates.

>>>>

---
title: "A Comprehensive Analysis Revealing Contributing Factors Behind Declining US Birth Rates"
#subtitle: "My subtitle if needed"
author: Maroosh Gillani, Aaron Liu & Jimmy Luc
thanks: "Code and data are available at: https://github.com/rex009x/us_birth_rates/tree/master/data. Selected aspects of the original paper that underpins this paper were reproduced and a record of this is available at: https://doi.org/10.48152/ssrp-nsc2-4v21 "
date: today
date-format: long
abstract: "The United States has been experiencing a steady decline in their birth rates, as suggested by Kearney, Levine, and Pardue's (2020) article 'The Puzzle of Falling US Birth Rates since the Great Recession'. Through the analysis of multiple different factors, such as age and education, we notice a continual decline in birth rate starting from the year 2007. We discover how different social, cultural, and economic shifts play a part in this worrying trend. This paper demonstrates how socioeconomic trends have affected Americans to shift their priorities, and how this change represented through data has played a key role in the downward trajectory of US birth rates."
format:
  pdf:
    fig-pos: "H"
toc: true
number-sections: true
bibliography: references.bib
---
```{r, include=FALSE, warning=FALSE, message=FALSE}
# workplace setup
library(tidyverse)
library(dplyr)
library(ggplot2)
library(here)
library(haven)

@@

The nation's birth rate has fallen significantly since 2007, dropping about 23% by 2022, as measured by the number of births per 1000 women [@citeAxios], and birth rates tend to fall based on the economy. 2007 marks the start of the most severe economic crisis the world has ever faced, spanning until 2009. The events of this crisis have had long lasting damage on the economy, with unemployment doubling and the crash of the housing market [@citeFDR]. These poor economic conditions and the unusually slow recovery of the economy would play a large role in the decision to have children, as an unpredictable economic environment makes people feel less secure about raising a family, thus contributing to the recorded decline in birth rates. This is just one of various factors that we will be exploring in this paper to observe any trends and correlations between societal and economic shifts and the dwindling birth rate in the US.

@@

The data for this paper was acquired through the replication package of Kearney, Levine, and Pardue's (2020) article "The Puzzle of Falling US Birth Rates since the Great Recession" available through the American Economic Association (AEA), with the aid of data available from Centers for Disease Control, and Prevention (CDC) Natality data [@citeCDCNCHS] and CDC SEER U.S. Population Data [@citeCDCSeer]. The data sources were accessed on February 12th, 2024 and act as the foundation of this paper. The data sources themselves are a collaborative effort including, but not limited to, the U.S. Census Bureau, the National Cancer Institute, SEER Data, CDC, and the National Center for Health Statistics (NCHS). The United States census is legally mandated by the United States Constitution which occurs every 10 years. The data consists largely of `numeric` values representing years, birth rates by race and ethnicity, marital status, education, age, and birth parity, or `character` values representing US states.

All data collection and analysis was done using statistical tools such as R [@citeR], and the data visualization program Rstudio IDE [@citeRStudio] to help streamline the workflow. The analysis was done using the R program and the following supporting packages `tidyverse` [@citeTidyverse], `haven` [@citeHaven], `ggplot2` [@citeGgplot2], `here` [@citeHere], `ggthemes` [@citeGgthemes], and `dplyr` [@citeDplyr]. We will provide more comprehensive details about the data cleaning, and analysis in the following subsections.

@@

The overarching variable in our measurements is birth rate. Birth rate is measured by the number of births per 1000 women. Using this metric, we graph it in the context of the other variables included in our research, as stated above.

@@

The data we acquired through various sources and methods, as previously mentioned, presents a plethora of data points which we concern ourselves with. In order to produce our desired visualizations, we must first begin by processing our data in preparation. While a portion of the data has already been processed by the authors of "The Puzzle of Falling US Birth Rates since the Great Recession" [@citeOriginalReplication], the remaining data sets are abundant. For example, there is data that is very specific, such as the population of women the age of 15 who live in Arkansas.

With consideration of our variables of interest in mind, we can proceed with the processing of the data. As birth rates are the main focal point of this paper, we are interested in the number of births that occur throughout the United States and the total population at any given time frame. The data including US Census Data and NCHS Vital Statistics Natality Birth Data provides this information by listing out the number of women and child births per state by year. Together, this provides us with sufficient information required to proceed with the visualization for the number of births per 1000 women in the US.

Next, we are interested in finding the number of births per 1000 women with regards to the woman's age. As previously mentioned, we have the required data points concerning the population of women, but now incorporating another variable of interest being the age. Then, we intend to organize the data into age groups in order to observe the potential generational trends in birth rate. Fortunately, the data is already prepared containing the birth rates per year for each corresponding five-year age group [@citeOriginalReplication].

@@

Considering births by "parity", referring to the number of children for a given woman, we are interested in the data containing information regarding the number of first, second, third, and fourth+ births in the US. Combining NCHS Natality Microdata and CDC SEER population data, we simply filter for the data points of interest to receive the information required to generate our visualizations.

Similar to the data processing to acquire the birth rates per year by five-year age groups, we gather and organize NCHS Vital Statistics birth data and CDC SEER population data by mother's age and birth cohorts. Data processing and visualization of @fig-3 was completed with the aid of Kimlin Chin's replication [@citeAnotherReplication] of Kearney, Levine, and Pardue's (2020) article "The Puzzle of Falling US Birth Rates since the Great Recession".

@@

The measurement of birth rates is nothing short of a difficult task considering all of its nuances. What exactly defines a birth and what parameters determine how it is recorded? This variable of interest becomes slightly cloudy when answering these questions. Do births that occur outside of a hospital count to the total number of births? The bounds of what a birth is becomes even blurrier when we consider miscarriages and related factors.

@@

Measuring the level of education appears to be simple but is more inconsistent than initially thought. Data is acquired through the CPS, which is a monthly voluntary survey where U.S. households are selected [@citeCPS]. This is important to note as there may be traces of response bias and other forms of biases. As a result, it is difficult to filter the data from these biased entries and the generated visualizations may be skewed.

The number of children for a given mother or "parity" is a rather straight-forward measurement. Although questions concerning whether legal adoption of a child contributes to this measurement remain, it is difficult to find an answer. A more concise explanation for the properties of this measurement would greatly benefit the reader and replicator's experience.

@@

The most likely explanation that acts as the root cause for the decline in US births is the shift in priorities across cohorts of women [@citeEconomicPolicy]. While there are many who can speculate and point fingers at specific events that occurred pre- or post-2007, such as the "inevitable" financial crisis more commonly known as "The Great Recession". In fact, @fig-1 implies a strong correlation between the 2007 crisis and declining birth rates, as it is clear to see that births per 1000 women have been on a sharp downwards trajectory since the aforementioned year.

Another factor to consider is that Fertility is pro-cyclical: there exists a relationship between the current economic state and fertility rate [@citeSageJournals]. This alone would suffice as the reasoning for such drop in fertility but instead the shifting priorities of young adults, adopting a new and modern take on life is the overarching explanation to the downward trajectory of US birth rates. This is supported by @fig-2a, as younger women, ranging from the ages of 15 to 29 years old have all exhibited decreasing births. @citeAgeArticle provides a possible explanation for this phenomenon. They state that younger women are more focused on their educations and careers, for the purpose of ensuring financial stability. Thus, they might feel unready to support a baby in their circumstances. In contrast, a comparatively older woman will have most likely become comfortable financially, and would feel ready to have, and support a baby.

@citeAgeArticle's point regarding education brings up some doubts regarding results seen in @fig-2b. Women pursuing higher education share similar birth rates over the years, all of which are significantly lower than women who do not have a high school degree. On one hand, one could justify this by considering the fact that women pursuing education likely have lesser time in their day, but at the same time, there are questions about how lesser educated women are able to financially afford taking care of their babies. @citeEducationArticle helps clear some of these doubts by pointing out that educated women have better access to birth control. Furthermore, they are also able to provide better care for their babies, resulting in an increased human capital for the baby, subsequently reducing the need for more children [@citeEducationArticle].

With @citeEducationArticle's information on why educated women have less children in mind, we look at the fact that female education in North America has experienced a steady rise from 1970 to 2020 [@citeNAFemEducation]. This can actually be seen in @fig-2c, which shows that lesser women have more than one child. A similar trend is seen in @fig-3, where the number of children ever born are declining at a significant rate as they are born to women belonging to more recent cohorts. This could be another explanation as to why birth rates are declining overall in the United States.
